---
title: 'Reproducible Research : Peer Assessment 1'
author: "Ishan Srivastava"
date: "Monday, May 11, 2015"
output: html_document
---
 
###Loading and Preprocessing Data 

```{r}
unzip("activity.zip")
data <- read.csv("activity.csv", colClasses = c("integer", "Date", "factor"))
data$month <- as.numeric(format(data$date, "%m"))
noNA <- na.omit(data)
rownames(noNA) <- 1:nrow(noNA)
head(noNA)
dim(noNA)
library(ggplot2)
```

##What is the mean total number of steps taken per day? 

```{r, echo = FALSE}
ggplot(noNA, aes(date, steps)) + geom_bar(stat = "identity", colour = "steelblue", fill = "steelblue", width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")
```

Mean total number of steps taken per day :- 

```{r}
totalsteps <- aggregate(noNA$steps, list(Date = noNA$date), FUN = "sum")$x
mean(totalsteps)
```

Median number of steps taken per day :- 

```{r}
median(totalsteps)
```

##What is the average daily activity pattern? 

```{r}
avgSteps <- aggregate(noNA$steps, list(interval = as.numeric(as.character(noNA$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"

ggplot(avgSteps, aes(interval, meanOfSteps)) + geom_line(color = "steelblue", size = 0.8) + labs(title = "Time Series Plot of the 5-minute Interval", x = "5-minute intervals", y = "Average Number of Steps Taken")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 

```{r}
avgSteps[avgSteps$meanOfSteps == max(avgSteps$meanOfSteps), ]
```

##Imputing Missing Values 

Total number of rows with NAs :- 
```{r}
sum(is.na(data))
```

```{r}
newData <- data
for (i in 1:nrow(newData)) {
  if(is.na(newData$steps[i])) {
    newData$steps[i] <- avgSteps[which(newData$interval[i] == avgSteps$interval), ]$meanOfSteps
  }
}

head(newData)
sum(is.na(newData))
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
ggplot(newData, aes(date, steps)) + geom_bar(stat = "identity", colour = "red", fill = "red", width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day (no missing data)", x = "Date", y = "Total number of steps")
```

Mean total number of steps taken per day :- 

```{r}
newTotalSteps <- aggregate(newData$steps, list(Date = newData$date), FUN = "sum")$x
newMean <- mean(newTotalSteps)
newMean
```

Median total number of steps taken per day :-

```{r}
newMedian <- median(newTotalSteps)
newMedian
```


So, after imputing the missing data, the new mean of total steps taken per day is the same as that of the old mean; the new median of total steps taken per day is greater than that of the old median.

##Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
head(newData)
newData$weekdays <- factor(format(newData$date, "%A"))
levels(newData$weekdays)
levels(newData$weekdays) <- list(weekday = c("Monday", "Tuesday", "Wednwsday", "Thursday", "Friday"), weekend = c("Saturday", "Sunday"))

levels(newData$weekdays)
table(newData$weekdays)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
avgSteps <- aggregate(newData$steps, list(interval = as.numeric(as.character(newData$interval)), weekdays = newData$weekdays), FUN = "mean")

names(avgSteps)[3] <- "meanOfSteps"
library(lattice)
xyplot(avgSteps$meanOfSteps ~ avgSteps$interval | avgSteps$weekdays, layout = c(1,2), type = "l", xlab = "Interval", ylab = "Number of Steps")
```

